Verifying a Postfix Language \\ \large CS595 Final Project

### 2021-12-01 project/paper.tex
day before presentation

\section{Syntax}
Taking the ideas from the above rant into account we arrive at something along these lines:

\renewcommand{\grammarlabel}[2]{#1 \hfill#2}
\begin{grammar}
    <$\tau$> ::= $\kwInt$
    |    $\kwLong$
    |    $\kwFloat$
    |    $\kwDouble$
    \alt $(\tau^*)$
    |    $\kwUnit$
    % |    $\tau_1~ \tau_2~ \texttt{|}$
    |    $\kwArrow{\tau_1} {\tau_2}$
    |    $\kwType{e}$

    <$\tau^*$> ::= $\varepsilon$ | $\tau^* ~ \tau$ | $\kwunpack{\tau}$

    <$e$> ::= $e~ e~ +$
    |    $e~ e~ \times$
    |    $\kwtern{e}{e}{e} $
    |    $e~ e~ ==$
    |    $\tau~ \tau~ ==$
    \alt $(e^*)$
    |    $\kwmacro{\tau^*}{e^*}$
    |    $\kwfix{e}{\tau}$
    \alt $\kwid{x}$
    |    $\kwid{T}$
    |    $e~ \textbf{@} $
    |    $e~ \kweid{x} =$

    <$e^*$> ::= $\varepsilon$ | $e^* ~ e$ | $\kwunpack{e}$
\end{grammar}
Notes
\begin{itemize}
    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
    \item Tuples and Tuple types are enclosed in parens
    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
\end{itemize}

\section{Semantics}
\subsection{Notation}
To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself. The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus can be omitted with this notation. The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding. This is explicitly defined in the syntax but implicitly expanded in the semantics.

@@

\subsection{Macros, Identifiers and Scoping}
The store, $\sigma$, consists of a list of \textit{scopes}, where each scope is a mapping from identifiers to values. Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope. A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure). Calling a macro value pushes a new empty mappings list onto the store, then places the macro's expressions and an $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}    
        % Closures capture lexically scoped variables via substitution 
        \Rule{LET}
            {v \val
                \\ \sigma = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n] ]
                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v] ]}
            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}

        \and \Rule{Identifier}
            {[ ... ~ \kwid{x} \step v ~ ... ] \in \sigma}
            {\sigma; \kwid{x} \step \sigma; v}

        \and \Rule{MACRO-2}
            {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n] e^*} \val}

        \and \Rule{CALL-1}
            {\sigma' = \sigma \cup [~] }
            {\sigma; \kwmacro{\tau^*}{e^*} \textbf{@} \step \sigma'; e^* \kw{end\_scope}}

@@

\subsection{Others}
Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}

        % unpack operator
        \Rule{TUPLES-3}
            {v_i \val ~ \forall~ i \in [1, n]}
            {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
        \Rule{TUPLES-4}
            {\strut}
            {\kwunpack{()} \step \varepsilon}
        \and

@@

\section{Semantics for Types}
Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types. These are provided by $\prim$ and $\tstep$ respectively.

@@

\subsection{Step Judgement}
Type expressions are evaluated left to right until everything is $\prim$.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of T* -> T*'
        \Rule{TSTEP}
            {\tau_1 \tstep \tau_1'
                \\\tau_i \prim ~ \forall \tau_i \in \tau^*}
            {\tau^* ~ \tau_1 ~ ... ~ \tau_n \tstep \tau^* ~ \tau_1' ~ ... ~ \tau_n}
        \and
    \end{mathpar}  
}

@@

        \and \Rule{Type-Id}
            {[ ... ~ \kwid{T} \tstep \tau ~ ... ] \in \sigma}
            {\sigma; \kwid{T} \tstep \sigma; \tau}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        % Tuple type literals evaluation
        \Rule{Tuple Types - 1}
            {\tau^* \tstep \tau^{*\prime}}
            {( \tau^* ) \tstep ( \tau^{*\prime} )}
        \Rule{Tuple Types - 2}
            {\tau_i \prim ~ \forall \tau_i \in \tau^*}
            {(\tau^*) \prim}
        \and

        % unpack operator
        \Rule{Tuple Types - 3}
            {\tau_i \prim ~ \forall~ i \in [1, n]}
            {\kwunpack{(\tau_1 ... \tau_n)} \tstep \tau_1 ... \tau_n}
        \Rule{Tuple Types - 4}
            {\strut}
            {\kwunpack{\kwUnit} \tstep \varepsilon}
        \and
    \end{mathpar}
}

@@

\section{Typing}
Note that because our notation is defined as ignoring irrelevant values there could be  
\subsection{Step Judgement}
{
    \centering
    \begin{mathpar}
    \Rule{T-0}
        {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
        {\typed{\ctx}{e}{\tau'}}
    \end{mathpar}
}

@@

\subsection{Macros}
This is why we needed the step judgement :sweatsmile:
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Move this to type rules   
        \Rule{Macro}
            {\tau^* = \tau_1 ~...~ \tau_n \\
             v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_n}{\tau_n} \\
             m = \kwmacro{\tau^*}{e^*} \\
             m \val }
            {\typed{\ctx}{m}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}

@@

        \and \Rule{Identifier}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}
        \and \Rule{Let}
            {\typed{\ctx}{e}{\tau}
                \\ \ctx(\kwid{x}) = e}
            {\typed{\ctx}{e \kwid{x}}{\varepsilon}}

        \and \Rule{Fix}
            {\tau_1 \prim
                \\ \typed{\ctx}{e}{\tau_2}
                \\ \tau_2 \prim
                \\ \tau_2 = \kwArrow{\tau^*_3 ~ \kwArrow{\tau^*_3}{\tau^*_4}}{\tau^*_4}}
            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\kwArrow{\tau^*_3}{\tau^*_4}}}
    \end{mathpar}
}

@@

            By induction, $e_1 \step e_1'$ or $e_1 \val$.
            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step e_1'~ e_2~ +$.

            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step e_1~ e_2'~ +$.

@@

            By induction, $e_1 \step e_1'$ or $e_1 \val$.
            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.

            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.

            If $e_1 \val$ and $e_2 \val$, then by induction, $e_3 \step e_3'$ or $e_3 \val$.
            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP-2}, $e \step \kwtern{e_1}{e_2}{e_3'}$.

@@

            % Then either all elements are values or there exists some *first* steppable element
            % if there's a steppable element, then e^*_t steps using STEP-{1,2}, so e^* steps using TUPLES-1
            % if there are no steppable elements, then all elements are values, so e^* \val by TUPLES-2
    \end{itemize}
\end{proof}

\subsection{Theorem - Preservation}
If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
\begin{proof}

\end{proof}

>>>>

\section{Syntax}
Here some operations also include their stack arguments, this is intended only for clarity as to what their function is as their arguments are come from the stack when it's their turn to be evaluated, not syntactically.

\renewcommand{\grammarlabel}[2]{#1 \hfill#2}
\begin{grammar}
    <$\tau$> ::= $\kwInt$
    |    $\kwLong$
    |    $\kwFloat$
    |    $\kwDouble$
    |    $(\tau^*)$
    |    $\kwUnit$
    % |    $\tau_1~ \tau_2~ \texttt{|}$
    |    $\kwArrow{\tau_1} {\tau_2}$
    \alt $\kwType{e}$
    |    $e~ \kweid{T} =$
    |    $\kwid{T}$
    |    $\kwunpack{\tau}$

    <$e$> ::= $e~ e~ +$
    |    $e~ e~ \times$
    |    $\kwtern{e}{e}{e} $
    |    $e~ e~ ==$
    |    $\tau~ \tau~ ==$
    \alt $(e^*)$
    |    $\kwmacro{\tau^*}{e^*}$
    |    $\kwfix{e}{\tau}$
    |    $\kwnumlit{n}{\kwInt}$
    |    $\kwnumlit{n}{\kwLong}$
    |    $\kwnumlit{n}{\kwFloat}$
    |    $\kwnumlit{n}{\kwDouble}$
    \alt $e~ \kweid{x} =$
    |    $\kwid{x}$
    |    $e~ \textbf{@} $
    |    $\kwunpack{e}$

    <$prog$> ::= $e^*$ | $\tau^*$ | $prog~ prog$ | $\varepsilon$
\end{grammar}
Notes
\begin{itemize}
    \item A program consists of a list of $op^*$ expressions
    \item $\varepsilon$ is used to denote an empty program/expression
    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
    \item Tuples and Tuple types are enclosed in parens
    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
\end{itemize}

\section{Small Step Semantics For Values}
\subsection{Notation}
To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself. The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus can be omitted with this notation. The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding. This is explicitly defined in the syntax but implicitly expanded in the semantics.

@@

\subsection{Macros, Identifiers and Scoping}
The store, $\sigma$, consists of a list of \textit{scopes}, where each scope is a mapping from identifiers to values. Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope. A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure). Calling a macro value pushes a new empty mappings list onto the store, then places the macro's expressions and an $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}    
        % Closures capture lexically scoped variables via substitution 
        \Rule{LET}
            {v \val
                \\ \sigma = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n] ]
                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n, x_{n+1} \step v] ]}
            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}

        \and \Rule{Identifier}
            { \kwid{x} \step v \in \sigma(-1)}
            {\sigma; \kwid{x} \step \sigma; v}

        \and \Rule{CALL-1}
            {\sigma' = \sigma \cup [~]
                \\ v = \kwmacro{\tau^*}{e^*}
                \\ v \val}
            {\sigma; v~ \textbf{@} \step \sigma'; e^* \kw{end\_scope}}

@@

\subsection{Others}
Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % unpack operator
        \Rule{TUPLES-3}
            {v_i \val ~\forall~ v_i \in v^*}
            {\kwunpack{(v^*)} \step v^*}

@@

\section{Small Step Semantics for Types}
Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types. These are provided by $\prim$ and $\tstep$ respectively.

@@

\subsection{Step Judgement}
Type expressions are evaluated left to right until everything is $\prim$. Types and expressions have equal precedence and both can be present on the stack.
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Derivation of T* -> T*'
        \Rule{TSTEP}
            {\tau \tstep \tau'
                \\\tau_i \prim ~ \forall~ \tau_i \in \tau_1^*}
            {\tau_1^* ~ \tau ~ \tau_2^* \tstep \tau_1^* ~ \tau' ~ \tau_2^*}

        \and \Rule{TSTEP - Mixed}
            {\tau^* \tstep \tau^{*\prime}
                \\ v_i \val ~ \forall v_i \in v^*}
            {v^*~ \tau^*~ e^*~ \tstep v^*~ \tau^{*\prime}~ e^*}
        \and \Rule{STEP - Mixed}
            {e^* \step e^{*\prime}
                \\ \tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\tau^*~ e^* ~ \tau_2^*  \step \tau^* ~ e^{*\prime} ~ \tau_1}
    \end{mathpar}
}

@@

        \and \Rule{MACRO-2}
            {\sigma = [ ..., [x_1 \step v_1 ~ ... ~ x_n \step v_n, 
                ~ T_1 \tstep \tau_1 ~ ... ~ T_n \tstep \tau_n ]]
                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n][\tau_1/T_1]...[\tau_n/T_n] e^*} \val}

        \and \Rule{Type-Id}
            { \kwid{T} \step \tau \in \sigma(-1)}
            {\sigma; \kwid{T} \tstep \sigma; \tau}
    \end{mathpar}
}

\subsection{Tuples}
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        % Tuple type literals evaluation
        \Rule{Tuple Types - 1}
            {\tau^* \tstep \tau^{*\prime}}
            {( \tau^* ) \tstep ( \tau^{*\prime} )}
        \Rule{Tuple Types - 2}
            {\tau_i \prim ~ \forall \tau_i \in \tau^*}
            {(\tau^*) \prim}
        \and
        % unpack operator
        \Rule{Tuple Types - 3}
            {\tau_i \prim ~ \forall~ \tau_i \in \tau^*}
            {\kwunpack{(\tau^*)} \tstep \tau_1 ... \tau_n}
    \end{mathpar}
}

@@

\section{Typing}
Note that we're using a somewhat weird notation for a stack machine (normally there's arrows representing stack transformation). This aligns with how we defined our syntax.

\subsection{Step Judgement}
{
    \centering
    \begin{mathpar}
    \Rule{T-0}
        {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
        {\typed{\ctx}{e}{\tau'}}
    \end{mathpar}
}

@@

\subsection{Macros}
This is why we needed the step judgement
{
    \centering
    \def \MathparLineskip {\lineskip=0.43cm}
    \begin{mathpar}
        % Move this to type rules   
        \Rule{Macro}
            {\tau^* = \tau_1 ~...~ \tau_n \\
             v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_n}{\tau_n} \\
             m = \kwmacro{\tau^*}{e^*} \\
             m \val }
            {\typed{\ctx}{\kwmacro{\tau^*}{e^*}}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}

@@

        \and \Rule{Identifier}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}
        \and \Rule{Let}
            {\typed{\ctx}{e}{\tau}
                \\ \ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{e~ \kwid{x} =}{\varepsilon}}

        \and \Rule{Fix}
            {\tau_1 \prim
                \\ \tau_1 = \kwArrow{(\tau^*_3)}{(\tau^*_4)}
                \\ \typed{\ctx}{e}{\kwArrow{(\tau^*_3 ~ \tau_1)}{(\tau^*_4)}}}
            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\tau_1}}
    \end{mathpar}
}

@@

            By induction, $e_1 \step e_1'$ or $e_1 \val$.
            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step e_1'~ e_2~ +$.

            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step e_1~ e_2'~ +$.

@@

            By induction, $e_1 \step e_1'$ or $e_1 \val$.
            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.

            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.

            If $e_1 \val$ and $e_2 \val$, then by induction, $e_3 \step e_3'$ or $e_3 \val$.
            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP}, $e \step \kwtern{e_1}{e_2}{e_3'}$.

@@

            If all elements are values, then $(e^*_t) \val$ by \Rulename{TUPLES-2}.

            Otherwise, one or more elements can step, including a leftmost steppable element $e_j \step e_j'$.
            Then, by \Rulename{STEP}, $e^*_t \step [e_j' / e_j] e^*_t$, so by \Rulename{Tuples-1}, $e^* \step ([e_j' / e_j] e^*_t)$.

        \item \Rulename{Unpack}
            Then $e^* = \kwunpack{(e^*_t)}$.
            By induction, either $(e^*_t) \step (e^{*'}_t)$ or $(e^*_t) \val$.

            If $(e^{*'}_t) \step (e^{*'}_t)$, then by \Rulename{STEP}, $\kwunpack{(e^*_t)} \step \kwunpack{(e^{*'}_t)}$.

            If $(e^*_t) \val$, then by \Rulename{TUPLES-3}, $\kwunpack{(e^*_t)} \step e^*_t$.

        \item \Rulename{Macro}
            Then $e^* = m = \kwmacro{\tau^*}{e^*_t}$. $m \val$ by the premise.

        \item \Rulename{Call}
            Then $e^* = e^*_x~ e_f~ \textbf{@}$. By induction, either $e_i \val~ \forall~ e_i \in e^*_x$ or $e^*_x \step e^{*'}_x$.

            If $e^*_x \step e^{*'}_x$, then by \Rulename{STEP}, $e^* \step e^{*'}_x~ e_f~ \textbf{@}$.

            Otherwise, $e_i \val~ \forall~ e_i \in e^*_x$ and, by induction, $e_f \val$ or $e_f \step e_f'$.

            If $e_f \step e_f'$, then by \Rulename{STEP}, $e^* \step e^*_x~ e_f'~ \textbf{@}$.

            Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.
    \end{itemize}
\end{proof}

\subsection{Theorem - Preservation}
If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
% \subsubsection{Lemma - Convergence}
% Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
% \begin{proof}
% Proof by exhaustion with all rules involving $\prim$ types not stepping. To values
% \end{proof}
% Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
% \begin{proof}
% This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does no have to follow recursive paths in order to get it's type (Type Rules \Rulename{Macro}, \Rulename{Fix}). The other rules are trivial (ie - step to a $\prim$) and/or irrelevant to the assertion
% \end{proof}

% \subsubsection{Proof}
% \begin{proof}
% \end{proof}
